---
title: "ARMA for Novice Flood Forecast Modellers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ARMA for Novice Flood Forecast Modellers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(reach.postproc)
render_mermaid <- function(diagram) {
  if (requireNamespace("DiagrammeR", quietly = TRUE)) return(DiagrammeR::mermaid(diagram))
  knitr::asis_output(paste0("```mermaid\n", diagram, "\n```"))
}
```

# What ARMA does

```{r flow, echo=FALSE}
render_mermaid("flowchart LR; A[Observation] --> C[Error]; B[Simulation] --> C; C --> D[AR model]; D --> E[Updated forecast]; B --> E")
```

AR projects the recent difference between observation and simulation. Positive error means the simulation is too low.

# Glossary

## AR
Autoregressive: predicts the next error from previous errors.

## MA
Moving average: represents new residual innovations.

## Characteristic root
A number describing decay, growth or oscillation.

## Decay time
How long a root remains important.

## Modal root
The EA convention. Stable modal roots lie inside the unit circle.

## Reciprocal lag root
The reciprocal convention. Stable roots lie outside the unit circle.

## Updated forecast
Simulation plus projected error.
