---
title: "Selecting a Default-Family Parameter Set"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Selecting a Default-Family Parameter Set}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 5.5,
  warning = FALSE,
  message = FALSE
)

library(reach.postproc)
library(data.table)
library(ggplot2)
```

# Selecting a default-family parameter set

A new AR model does not require three independently estimated decay times. In the default family, the principal root controls most long-duration behaviour. The middle and rapid roots shape the response shortly after the forecast starts. It is therefore sufficient to select one principal timescale for the catchment.

Suitable evidence includes catchment response time, FEH lag, event analysis and a broad rapid, medium or slow response classification. The five-day IMFS forecast horizon and the inability of fixed-parameter AR to represent different dry- and wet-period timescales mean that the selected value should be treated as a useful catchment classification, not a precisely calibrated physical constant.

## Continuous analytical construction

```{r default-family-continuous}
parameters <- default_family_ar_parameters(
  principal_decay = 9,
  units = "hours",
  time_step_minutes = 15
)

parameters@coefficients
root_table(roots(parameters, time_step_minutes = 15))
```

This calculation does not interpolate between the six-hour and twelve-hour rows of a table. It sets the nine-hour principal root directly, solves the middle root while retaining the default-family constraints and converts the three roots to AR coefficients.

## Standard published choices

Use a named standard set where configuration consistency and easy recognition are priorities.

```{r default-family-standard}
parameters <- standard_family_ar_parameters("12 hours")
parameters@coefficients
```

The full catalogue is available programmatically:

```{r default-family-table}
standard_family_ar_table()
```

The table is useful for governance, recognition and validation. It is not required for calculation and should not be used for linear interpolation of coefficients.

## Precision

Small changes in the second and third coefficients can produce large changes in principal decay time. Keep full numerical precision within R. Where values must be transferred manually, retain at least five significant figures and reassess the resulting roots after transfer.

## Mathematical construction

The standard table contains selected points from a continuous analytical family. The coefficients are not obtained by interpolating between adjacent rows.

For a principal decay time $\tau_1$, expressed in model timesteps, the principal modal root is

$$
z_1 = \exp\left(-\frac{1}{\tau_1}\right).
$$

For an infinite principal decay time,

$$
z_1 = 1.
$$

The rapid root is defined by a decay time $\tau_3$ and oscillation period $T_3$:

$$
z_3 = \exp\left(-\frac{1}{\tau_3}\right)
      \exp\left(\frac{2\pi i}{T_3}\right).
$$

The default family uses

$$
\tau_3 = \frac{1}{3},
\qquad
T_3 = 2.
$$

Because $\exp(\pi i)=-1$, the rapid root is a negative real root:

$$
z_3 = -\exp(-3) \approx -0.04978707.
$$

For a third-order model, the first AR coefficient is the sum of the three roots:

$$
a_1 = z_1 + z_2 + z_3.
$$

The default family fixes

$$
a_1 = 1.765.
$$

The middle positive root is therefore solved directly:

$$
z_2 = 1.765 - z_1 - z_3.
$$

Its decay time is

$$
\tau_2 = -\frac{1}{\ln(z_2)}.
$$

Once the three modal roots are known, the characteristic polynomial is

$$
(z-z_1)(z-z_2)(z-z_3)
=
z^3-a_1z^2-a_2z-a_3.
$$

Expanding the product gives the Deltares AR coefficients:

$$
a_1 = z_1+z_2+z_3,
$$

$$
a_2 = -(z_1z_2+z_1z_3+z_2z_3),
$$

and

$$
a_3 = z_1z_2z_3.
$$

This is the calculation performed by `default_family_ar_parameters()`. The only empirical decision is the choice of principal decay time. The conversion from that timescale to roots and coefficients is deterministic.

### Worked nine-hour example

For a 15-minute model timestep, a nine-hour principal decay corresponds to

$$
\tau_1 = \frac{9\times60}{15}=36
$$

timesteps. The principal root is therefore

$$
z_1 = \exp\left(-\frac{1}{36}\right).
$$

The package then calculates $z_2$, retains the fixed rapid root $z_3$, and converts the three roots to coefficients.

```{r default-family-nine-hour-maths}
nine_hour <- default_family_ar_parameters(
  principal_decay = 9,
  units = "hours",
  time_step_minutes = 15
)

attr(nine_hour, "modal_roots")
attr(nine_hour, "middle_decay_steps")
nine_hour@coefficients
```

### Why the coefficients change so little

The root modulus is related exponentially to decay time:

$$
|z|=\exp\left(-\frac{1}{\tau}\right).
$$

As $\tau$ becomes large, $|z|$ approaches one. Large changes in a long decay time therefore correspond to small changes in the principal root and still smaller changes in the resulting coefficients. This explains why parameter sets with markedly different principal decay times can have visually similar values of $a_2$ and $a_3$.

It also explains why coefficient rounding matters. A small numerical change in a coefficient can imply a much larger change in the longest decay time. Calculate and store coefficients at full precision; round only for display.
