---
title: "Mathematical and Developer Validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mathematical and Developer Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 5.5,
  warning = FALSE,
  message = FALSE
)

library(reach.postproc)
library(data.table)
library(ggplot2)

render_mermaid <- function(diagram) {
  if (requireNamespace("DiagrammeR", quietly = TRUE)) {
    return(DiagrammeR::mermaid(diagram))
  }

  knitr::asis_output(
    paste0("```mermaid\n", diagram, "\n```")
  )
}
```


# Purpose

This article records the mathematical contracts and numerical equivalences that protect the package from subtle sign, root and indexing errors.

```{r validation-flow, echo=FALSE}
render_mermaid("
flowchart LR
    A[Parameters] --> B[Modal roots] --> C[Reconstructed parameters]
    D[AR recurrence] --> E[Modal decomposition]
    F[Fixed-lead recurrence] --> G[Fixed-lead root projection]
    H[ARMA recurrence] --> I[Impulse response]
")
```

# Characteristic polynomial

For Deltares coefficients,

$$
z^p-a_1z^{p-1}-a_2z^{p-2}-\cdots-a_p=0.
$$

The modal solution is

$$
x_t=\sum_{n=1}^{p}c_nz_n^t.
$$

# Root-to-parameter conversion

For modal root $r$, the polynomial factor is

$$
z-r.
$$

When polynomial coefficients are stored in ascending order, the factor is

```r
c(-r, 1)
```

Using `c(1, -r)` would create $1-rz$ and therefore introduce reciprocal roots.

```{r developer-roundtrip}
p1 <- default_ar_parameters()
z1 <- roots(p1)@values
p2 <- roots_to_parameters(z1)
z2 <- roots(p2)@values

p1@coefficients
p2@coefficients
```

The test suite should order complex roots deterministically before comparing them, because `polyroot()` does not guarantee a semantic root order.

# Raw and displayed decay time

The mathematical decay parameter is

$$
\tau=-\frac{1}{\ln|z|}.
$$

Growing roots have negative raw $\tau$. Operational diagnostic tables may display infinite decay for growth, but should retain the raw value separately.

# Modal decomposition

The root weights solve the recent errors at times

$$
-1,-2,\ldots,-p.
$$

Modal time zero is therefore the first forecast value. Package indexing is:

```text
modal time 0 = forecast step 1
modal time 1 = forecast step 2
modal time 2 = forecast step 3
```

```{r developer-decomposition}
parameters <- default_ar_parameters()
forecast <- forecast_ar(
  parameters,
  initial_errors = c(0.20, 0.15, 0.10),
  steps = 100L
)

components <- decompose_ar(
  parameters,
  initial_errors = c(0.20, 0.15, 0.10),
  steps = 100L
)

reconstructed <- components[
  ,
  .(value = sum(contribution_real)),
  by = step
]

all.equal(
  reconstructed$value,
  forecast_series(forecast)$ar_error,
  tolerance = 1e-9
)
```

# Fixed-lead equivalence

The recurrence implementation is the production path. Independent root projection is a valuable cross-check.

For target index $i$ and lead $L$, initialise from errors available at $i-L$, then project exactly $L$ steps.

The test suite should compare recurrence and root projection for multiple leads, error histories, real roots and complex conjugate roots.

# ARMA response equivalence

For deterministic residual sequence $\varepsilon_t$, the recurrence is

$$
x_t=\sum_{n=1}^{p}a_nx_{t-n}
+\varepsilon_t
+\sum_{m=1}^{q}b_m\varepsilon_{t-m}.
$$

A unit impulse should produce the same series through direct recurrence and the response interface.

# Numerical conditioning

Root decomposition solves a Vandermonde-like system. Closely spaced roots can cause poor conditioning. Development checks should inspect the condition number and warn where the decomposition may be numerically fragile.

# Assessment boundaries

Tests should cover values immediately below, on and above each boundary:

- modulus one;
- decay time 240;
- decay time 8;
- rapid-decay exception at one;
- oscillation ratio 4.605;
- permitted and non-permitted model orders.

# Sign-convention invariance

Deltares and standard inputs that differ only by sign convention must produce identical internal coefficients, roots and forecasts.

# Plot contracts

The unit-circle polygon should:

- close exactly;
- have modulus one at every coordinate;
- use fixed aspect ratio;
- display stable modal roots inside;
- display stable reciprocal roots outside;
- expand limits for unstable or large reciprocal roots.

# Reproducibility contract

Public calculation functions should:

- avoid hidden file output;
- avoid changing caller-owned tables by reference;
- return stable column names;
- retain units and identifiers as metadata;
- expose diagnostics rather than silently repair data;
- fail early for impossible inputs.

# Minimum developer checks

```r
# Not run in the vignette build.
devtools::document()
devtools::load_all()
devtools::test()
devtools::build_vignettes()
devtools::check()
```

The package is ready for release only when the numerical tests and package checks pass together.
