---
title: "Mathematical and Developer Validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mathematical and Developer Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 5.5,
  warning = FALSE,
  message = FALSE
)

library(reach.postproc)
library(data.table)
library(ggplot2)

render_mermaid <- function(diagram) {
  escaped_diagram <- htmltools::htmlEscape(
    diagram
  )

  knitr::asis_output(
    paste0(
      '<pre class="mermaid">',
      escaped_diagram,
      '</pre>'
    )
  )
}
```


# Purpose

This article records the mathematical contracts and numerical equivalences that protect the package from subtle sign, root and indexing errors.

```{r validation-flow, echo=FALSE}
render_mermaid("
flowchart LR
    A[Parameters] --> B[Modal roots] --> C[Reconstructed parameters]
    D[AR recurrence] --> E[Modal decomposition]
    F[Fixed-lead recurrence] --> G[Fixed-lead root projection]
    H[ARMA recurrence] --> I[Impulse response]
")
```

# Characteristic polynomial

For Deltares coefficients,

$$
z^p-a_1z^{p-1}-a_2z^{p-2}-\cdots-a_p=0.
$$

The modal solution is

$$
x_t=\sum_{n=1}^{p}c_nz_n^t.
$$

# Modal roots versus reciprocal lag roots

The polynomial above is solved directly for $z$ in $x_t=c\,z^t$. Call these the **modal roots**. Stable modal roots lie inside the unit circle. `roots()`, `assess()` and `plot_ar()`'s default view all use this convention.

A different, equally standard construction exists. Write the recurrence with the backshift operator $B$ (where $Bx_t=x_{t-1}$) and solve $\phi(B)=0$ directly for $B$:

$$
1-a_1B-a_2B^2-\cdots-a_pB^p=0.
$$

This is the Box-Jenkins/ARIMA textbook convention, and its stable region is *outside* the unit circle. The two polynomials are reciprocals of one another: substituting $B=1/z$ into the modal polynomial and clearing denominators recovers the backshift polynomial, so each backshift root is the reciprocal of the matching modal root.

```{r modal-vs-lag}
parameters <- ar_parameters(c(2.1, -1.6, 0.4))

# Modal roots: direct solution of z^3 - 2.1z^2 + 1.6z - 0.4 = 0
modal <- polyroot(c(-0.4, 1.6, -2.1, 1))

# Reciprocal lag (backshift/Box-Jenkins) roots: 1 - 2.1z + 1.6z^2 - 0.4z^3 = 0
lag <- polyroot(c(1, -2.1, 1.6, -0.4))

modal
lag

all.equal(sort(Mod(1 / modal)), sort(Mod(lag)), tolerance = 1e-9)
```

`root_table()` reports both without requiring either polynomial to be built by hand: `root_real`/`root_imaginary` for the modal form, `lag_root_real`/`lag_root_imaginary` for the reciprocal.

```{r modal-vs-lag-table}
root_table(roots(parameters))[
  ,
  .(root_real, root_imaginary, lag_root_real, lag_root_imaginary)
]
```

An external check of this package's root calculations that assumes the Box-Jenkins convention will report different root values, and a different stability region (outside rather than inside the unit circle), to those `reach.postproc` reports. That is not a disagreement about the arithmetic. It is two valid conventions describing the same model. Confirm which convention a comparison is using before treating a mismatch as a defect.

# Floating-point residuals on real roots

A root that is mathematically real, such as the `0.5` above, will rarely print as a clean `0.5+0i`. Expect something like `0.5+8.414707e-16i` instead. That residual is floating-point rounding noise, not a genuine imaginary component, and it is not a sign that anything has gone wrong.

Two different routes to the same true value accumulate different, unrelated rounding dust, because they perform different sequences of floating-point operations to get there:

```{r floating-point-residual}
modal_roots <- polyroot(c(-0.4, 1.6, -2.1, 1))
inverted_roots <- 1 / polyroot(c(1, -2.1, 1.6, -0.4))

modal_roots[order(Re(modal_roots), Im(modal_roots))]
inverted_roots[order(Re(inverted_roots), Im(inverted_roots))]
```

Both real roots sit at `0.5`. Their residual imaginary parts are both around $10^{-15}$ to $10^{-16}$, which is the scale of `.Machine$double.eps` (roughly $2.2\times10^{-16}$) -- the smallest relative difference a double-precision number can represent. They are not equal to each other bit for bit, because `modal_roots` came from one `polyroot()` call solving one cubic, and `inverted_roots` came from a *different* `polyroot()` call solving a different cubic, followed by a division. Different arithmetic, same true answer, different leftover noise. This happens on any platform, with any sequence of floating-point operations that arrive at the same mathematical result by different paths; it is a property of finite-precision arithmetic, not of `polyroot()`, R, or this package.

This is exactly why nothing in `reach.postproc` compares roots with exact equality. `roots()`'s own real/complex classification,

```r
real <- abs(Im(z)) <= tolerance
```

uses `tolerance = sqrt(.Machine$double.eps)` ($\approx 1.49\times10^{-8}$) by default -- several orders of magnitude larger than any floating-point residual this kind of calculation produces, so genuine noise is always classified as real. `roots_to_parameters()`'s conjugate-pair check and the `all.equal(..., tolerance = ...)` calls used for verification throughout this vignette follow the same principle. A residual at $10^{-15}$ or $10^{-16}$ is never worth investigating. A residual anywhere near $10^{-8}$ or larger is a different matter, and should be.

# Root-to-parameter conversion

For modal root $r$, the polynomial factor is

$$
z-r.
$$

When polynomial coefficients are stored in ascending order, the factor is

```r
c(-r, 1)
```

Using `c(1, -r)` would create $1-rz$ and therefore introduce reciprocal roots.

```{r developer-roundtrip}
p1 <- default_ar_parameters()
z1 <- roots(p1)@values
p2 <- roots_to_parameters(z1)
z2 <- roots(p2)@values

p1@coefficients
p2@coefficients
```

The test suite should order complex roots deterministically before comparing them, because `polyroot()` does not guarantee a semantic root order.

# Raw and displayed decay time

The mathematical decay parameter is

$$
\tau=-\frac{1}{\ln|z|}.
$$

Growing roots have negative raw $\tau$. Operational diagnostic tables may display infinite decay for growth, but should retain the raw value separately.

# Modal decomposition

The root weights solve the recent errors at times

$$
-1,-2,\ldots,-p.
$$

Modal time zero is therefore the first forecast value. Package indexing is:

```text
modal time 0 = forecast step 1
modal time 1 = forecast step 2
modal time 2 = forecast step 3
```

```{r developer-decomposition}
parameters <- default_ar_parameters()
forecast <- forecast_ar(
  parameters,
  initial_errors = c(0.20, 0.15, 0.10),
  steps = 100L
)

components <- decompose_ar(
  parameters,
  initial_errors = c(0.20, 0.15, 0.10),
  steps = 100L
)

reconstructed <- components[
  ,
  .(value = sum(contribution_real)),
  by = step
]

all.equal(
  reconstructed$value,
  forecast_series(forecast)$ar_error,
  tolerance = 1e-9
)
```

# Fixed-lead equivalence

The recurrence implementation is the production path. Independent root projection is a valuable cross-check.

For target index $i$ and lead $L$, initialise from errors available at $i-L$, then project exactly $L$ steps.

The test suite should compare recurrence and root projection for multiple leads, error histories, real roots and complex conjugate roots.

# ARMA response equivalence

For deterministic residual sequence $\varepsilon_t$, the recurrence is

$$
x_t=\sum_{n=1}^{p}a_nx_{t-n}
+\varepsilon_t
+\sum_{m=1}^{q}b_m\varepsilon_{t-m}.
$$

A unit impulse should produce the same series through direct recurrence and the response interface.

# Numerical conditioning

Root decomposition solves a Vandermonde-like system. Closely spaced roots can cause poor conditioning. Development checks should inspect the condition number and warn where the decomposition may be numerically fragile.

# Assessment boundaries

Tests should cover values immediately below, on and above each boundary:

- modulus one;
- decay time 240;
- decay time 8;
- rapid-decay exception at one;
- oscillation ratio 4.605;
- permitted and non-permitted model orders.

# Sign-convention invariance

Deltares and standard inputs that differ only by sign convention must produce identical internal coefficients, roots and forecasts.

# Plot contracts

The unit-circle polygon should:

- close exactly;
- have modulus one at every coordinate;
- use fixed aspect ratio;
- display stable modal roots inside;
- display stable reciprocal roots outside;
- expand limits for unstable or large reciprocal roots.

# Reproducibility contract

Public calculation functions should:

- avoid hidden file output;
- avoid changing caller-owned tables by reference;
- return stable column names;
- retain units and identifiers as metadata;
- expose diagnostics rather than silently repair data;
- fail early for impossible inputs.

# Minimum developer checks

```r
# Not run in the vignette build.
devtools::document()
devtools::load_all()
devtools::test()
devtools::build_vignettes()
devtools::check()
```

The package is ready for release only when the numerical tests and package checks pass together.
