---
title: "Event Triggered and Time Jumped AR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Event Triggered and Time Jumped AR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE,comment="#>",fig.width=9,fig.height=5.5)
library(reach.postproc)
library(data.table)
library(ggplot2)
```

# Status and scope

Event Triggered AR (ET-AR) and Time Jumped AR (TJ-AR) were developed in 2025–2026 to address limitations of fixed-parameter AR updating without introducing a high-order model. The package marks trigger choice and CWI adjustment as experimental because site-independent trigger-selection procedures have not yet been established.

# Event Triggered AR

ET-AR maintains two parameter sets. The steady set has a long principal decay time. The event set has a principal decay time related to catchment response. Forecasts start in the steady state and switch once to the event state when a trigger is met.

```{r et-parameters}
steady <- default_et_ar_steady_parameters()
event <- event_ar_parameters(response_time_steps = 48)
root_table(roots(steady, time_step_minutes = 15))
root_table(roots(event, time_step_minutes = 15))
```

The recurrence state is retained at the switch, so there is no discontinuous reset of the error amplitude.

## Rainfall accumulation trigger

```{r rainfall-trigger}
rainfall <- c(rep(0,8),1,2,3,4,2,1,rep(0,34))
trigger <- rainfall_accumulation_trigger(rainfall,window_steps=8L,threshold=10)
config <- et_ar_configuration(steady,event,trigger,metadata=list(response_time_source="site assessment"))
simulation <- 0.7 + 1.2*exp(-((seq_along(rainfall)-25)/7)^2)
result <- forecast_et_ar(config,c(.2,.19,.18),simulation,time_step_minutes=15)
series <- et_ar_series(result)
ggplot(series,aes(lead_time_minutes/60,updated,colour=parameter_state))+geom_line()+geom_vline(xintercept=series$lead_time_minutes[result$switch_step]/60,linetype="dashed")+theme_minimal()+labs(x="Lead time, hours",y="Updated value",colour="Parameters")
```

## Updated level or flow trigger

```{r threshold-trigger}
threshold_config <- et_ar_configuration(
  steady,event,
  updated_threshold_trigger(threshold=1.0,initial_value=.8,quantity="level")
)
threshold_result <- forecast_et_ar(threshold_config,c(.2,.19,.18),simulation)
et_ar_series(threshold_result)
```

The threshold is evaluated against the previous updated value, not the simulation alone.

## CWI-adjusted rainfall trigger

The experimental factor is one at or below CWI 125, declines linearly to zero at CWI 165 and is zero thereafter.

```{r cwi-trigger}
cwi_rainfall_factor(c(120,125,145,165,170))
cwi_trigger <- cwi_adjusted_rainfall_trigger(rainfall,cwi=145,dry_threshold=20,window_steps=8L)
```

# Time Jumped AR

TJ-AR selects `p` historic errors separated by jump `J` rather than using consecutive points. This can reduce sensitivity to short-period noise where `J` is greater than the noise decorrelation time.

```{r tj-example}
set.seed(42)
history <- seq(.25,.08,length.out=40)+rnorm(40,0,.015)
tj <- forecast_tj_ar(default_ar_parameters(),history,jump=5L,steps=96L)
tj$selected_inputs
tj_ar_series(tj)
```

`jump = 1` reproduces conventional AR.

```{r jump-comparison}
jump_plot <- plot_jump_sensitivity(
  default_ar_parameters(),
  history,
  jumps = c(
    1L,
    3L,
    5L
  ),
  steps = 96L
)

jump_plot

attr(
  jump_plot,
  "failed_jumps"
)
assess_jump_size(default_ar_parameters(),jump=5L,noise_decorrelation_steps=3)
```

A large jump can suppress the fast root and eventually the middle root. The method can then resemble AR(2) or AR(1). The diagnostic reports this risk without imposing a universal threshold.

# Combined ET-TJ-AR

TJ-AR first reconstructs the recent consecutive state. ET-AR then advances the normal recurrence and switches parameters when required.

```{r combined}
combined <- forecast_et_tj_ar(config,history,jump=5L,simulated=simulation)
combined$tj_initialisation$selected_inputs
et_ar_series(combined)
```

# Operational assurance

Before operational use, document:

- steady and event parameter provenance;
- catchment response-time source;
- trigger type, window and threshold;
- jump-size evidence;
- input domain and units;
- rating and threshold versions;
- behaviour over several events, start times and antecedent conditions;
- effects on downstream model boundaries.

# Terminology

**ET-AR** switches once from steady to event parameters. **TJ-AR** uses non-consecutive errors to estimate modal weights. **Jump** is the timestep separation between selected errors. **CWI** is Catchment Wetness Index. **DMCA** is Detrending Moving-average Cross-correlation Analysis.
