---
title: "Characteristic Roots for Novice Flood Forecast Modellers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Characteristic Roots for Novice Flood Forecast Modellers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 5.5,
  warning = FALSE,
  message = FALSE
)

library(reach.postproc)
library(data.table)
library(ggplot2)

render_mermaid <- function(diagram) {
  escaped_diagram <- htmltools::htmlEscape(
    diagram
  )

  knitr::asis_output(
    paste0(
      '<pre class="mermaid">',
      escaped_diagram,
      '</pre>'
    )
  )
}
```


# Purpose

["ARMA for Novice Flood Forecast Modellers"](arma-for-novices.html) explains what AR does and why. This vignette is the deep dive on one part of that picture: characteristic roots, what they are, how `reach.postproc` uses them, and what they mean for whether a parameter set is safe to run operationally. No background in time-series mathematics is assumed.

# What a root actually is

The error between an observation and a model's simulation rarely moves in one simple way. It can fade quietly, linger stubbornly, or flip between positive and negative a few times before settling down. A root describes **one** of those behaviours in isolation.

Think of it like this: the projected error isn't one wobble, it's several wobbles happening at once, added together. A third-order (AR(3)) model has three roots, so three wobbles. One might be a slow, gentle fade. Another might fade twice as fast. A third might flip sign at every timestep while shrinking rapidly. Add all three together and you get the projected error the package actually reports.

```{r wobbles-flow, echo=FALSE}
render_mermaid("
flowchart LR
    A[Root 1: a slow fading wobble]
    B[Root 2: a faster fading wobble]
    C[Root 3: a quick flip-flopping wobble]
    D[Add them together]
    E[Projected error]

    A --> D
    B --> D
    C --> D
    D --> E
")
```

You never have to work out the wobbles yourself. [`roots()`](../reference/roots.html) does that for you, starting from the AR coefficients you supply. What matters is being able to read what comes back, because that's what tells you whether a parameter set will behave sensibly before you ever run it against real data.

# Reading one root

A root is a single point on a two-dimensional map: it has a **real** part and (sometimes) an **imaginary** part. Where that point sits tells you everything about how its wobble behaves.

The reference point is the **unit circle** — an imaginary circle of radius one drawn around zero. `reach.postproc` always plots this circle alongside the roots, because where a point falls relative to it is the whole story:

```{r stability-flow, echo=FALSE}
render_mermaid("
flowchart TD
    A{Where does the root sit?}
    B[Inside the unit circle]
    C[On the unit circle]
    D[Outside the unit circle]
    E[Fades away: stable]
    F[Never fades: marginal]
    G[Grows forever: unstable]

    A --> B --> E
    A --> C --> F
    A --> D --> G
")
```

A root's distance from zero is called its **modulus**. Modulus less than one means stable (the wobble fades). Modulus exactly one means marginal (it never fades, it never grows, it just sits there forever). Modulus greater than one means unstable (the wobble grows, which is never acceptable in an operational parameter set).

Try it on the Environment Agency's own defaults:

```{r roots-novice-defaults}
parameters <- default_ar_parameters()
root_information <- roots(parameters)

root_table(root_information)[, .(root_id, root_real, root_imaginary, root_modulus, modal_stability)]
plot_ar(root_information, root_definition = "modal")
```

All three points sit inside the circle, and the `modal_stability` column confirms it: `Stable`. That's a basic precondition for a usable parameter set, though not the only one — more on that below.

# Decay time

Knowing a root is stable tells you *that* it fades. Decay time tells you *how fast*. It's the number of timesteps it takes for a non-oscillating wobble to fall to about a third of its starting size (37%, to be precise — the same "about a third" rule that governs anything decaying exponentially).

```{r roots-novice-decay}
root_table(root_information)[, .(root_id, decay_time_steps, decay_time_hours)]
```

A root with a decay time of 2 steps has essentially vanished within a handful of timesteps: useful for reacting to a sudden change, useless for remembering anything beyond the immediate present. A root with a decay time of 200 steps barely moves across an entire forecast: useful for representing a slowly-drifting bias, dangerous if it's slow enough to keep an old, now-irrelevant error alive for days. `reach.postproc`'s [`assess()`](../reference/assess.html) checks both ends of that range, which the next section covers.

# Oscillation and complex roots

Some wobbles don't just shrink smoothly, they swing between positive and negative while they shrink — like a plucked guitar string that's getting quieter but still vibrating back and forth. A root like this isn't one number, it's a **pair**: two complex roots with matching real parts and opposite imaginary parts, called a **conjugate pair**. You never see half a swinging wobble on its own; the maths requires both halves of the pair to exist together, or the result wouldn't make physical sense.

The Environment Agency's own defaults already contain one oscillating root — notice `root_id` 3 above has a nonzero `root_imaginary`. It's a real number that happens to be negative, which produces the simplest possible oscillation: a flip in sign at every single timestep. That's one edge case of oscillation. The more general case is a true complex-conjugate pair, swinging with a longer period. Here's a parameter set built to show one:

```{r roots-novice-oscillating}
oscillating <- ar_parameters_from_timescales(
  decay_times = 20,
  oscillation_periods = 8,
  label = "Oscillating pair example"
)

oscillating@order

oscillating_roots <- roots(oscillating)
root_table(oscillating_roots)[, .(root_id, root_real, root_imaginary, root_modulus, oscillation_period_steps)]
plot_ar(oscillating_roots, root_definition = "modal")
```

Two roots, same real part, opposite imaginary parts — a genuine conjugate pair, visibly off the horizontal axis in the plot. The `oscillation_period_steps` column says this pair completes one full swing every 8 timesteps. A single `decay_times`/`oscillation_periods` entry like this always produces **two** roots once its conjugate partner is added automatically, so a model order can be larger than the number of timescales you asked for — worth remembering when you're counting up to AR(2) or AR(3).

Compare that against a root with `oscillation_period_steps` of `Inf`: that's not oscillating at all, it's a plain positive real root settling smoothly. `Inf` here doesn't mean "forever growing," it means "no oscillation, ever" — a common point of confusion worth flagging explicitly.

# How the package uses roots

Everything downstream of a parameter set flows from its roots:

```{r roots-novice-pipeline, echo=FALSE}
render_mermaid("
flowchart LR
    A[AR coefficients]
    B[roots]
    C[root_table and plot_ar]
    D[assess]
    E[decompose_ar]
    F[Pass or fail decision]
    G[See each wobble separately]

    A --> B
    B --> C
    B --> D --> F
    B --> E --> G
")
```

- [`roots()`](../reference/roots.html) calculates the roots once and packages them up for everything else to use.
- [`root_table()`](../reference/root_table.html) and [`plot_ar()`](../reference/plot_ar.html) let you inspect them directly, as above.
- [`assess()`](../reference/assess.html) runs the Environment Agency's acceptance criteria against the roots and returns a pass or fail with a plain-English summary.
- [`decompose_ar()`](../reference/decompose_ar.html) shows each root's individual wobble over time, rather than only the sum — useful when a projected error looks odd and you want to know which root is responsible.

Further into the package, the event-triggered and time-jumped AR methods (`assess_jump_size()`, `forecast_tj_ar()`) rank roots by decay time to judge whether a chosen "jump" size is sensible. If that territory is new to you, see the ["Event Triggered and Time Jumped AR"](event-triggered-and-time-jumped-ar.html) vignette once you're comfortable with the material here — the same unit-circle thinking carries straight over.

# What stability means for your forecast

"Stable" is not a maths nicety, it's an operational requirement. Translating the acceptance criteria `assess()` applies:

- **Any growing root fails outright.** A correction that gets bigger every timestep, forever, is never acceptable in a flood forecast.
- **A root that decays too slowly fails.** If a root barely fades across the whole forecast horizon, a single bad observation at the start can keep distorting the update long after it should have been forgotten.
- **A model where every root decays too fast fails.** If nothing sticks around for more than a step or two, the AR model isn't using any real memory of the recent error, and arguably isn't doing anything worth having.
- **Strong, slow oscillation fails.** A correction that keeps flipping sign every few steps looks like noise, undermines confidence in the forecast, and rarely reflects anything physically real. A very rapid oscillation is tolerated only when it also disappears almost immediately, as with the EA defaults' fast root above.
- **Only AR(2) and AR(3) are permitted by default.** Higher orders aren't inherently invalid mathematically, but they fall outside what's been operationally reviewed.

```{r roots-novice-assessment}
assessment <- assess(parameters)
assessment@summary
assessment_table(assessment)
```

A pass means the parameter set clears these structural checks. It does not mean every update will improve on the raw simulation — that depends on the actual recent error history, which varies event by event. Checking roots tells you the model *can* behave sensibly; checking performance against real events (see ["Using reach.postproc"](using-reach-postproc.html)) tells you whether it *did*.

# Try it yourself

Three small, hand-checkable examples, if you want to build your own intuition before trusting the plots:

```{r roots-novice-try-1}
# A simple AR(2) with a genuine complex pair: roots at 0.8 +/- 0.4i
ar_parameters(c(1.6, -0.8)) |> roots() |> root_table()
```

```{r roots-novice-try-2}
# A second AR(2), a different angle: roots at 0.5 +/- 0.5i
ar_parameters(c(1, -0.5)) |> roots() |> root_table()
```

```{r roots-novice-try-3}
# AR(3): one real root at 0.5, plus the same complex pair as the first example
ar_parameters(c(2.1, -1.6, 0.4)) |> roots() |> root_table()
```

Each one is small enough to factorise and check by hand if you want to be completely sure the package agrees with you, not just the other way round.

# Key messages

1. A root describes one wobble within the projected error; the package adds them together.
2. Modulus tells you whether a wobble fades (inside the unit circle), persists forever (on it), or grows (outside it) — only "inside" is acceptable.
3. Decay time tells you how fast a wobble fades; both too fast and too slow can fail assessment.
4. Complex roots always come in conjugate pairs and describe genuine oscillation; `oscillation_period_steps` of `Inf` means no oscillation at all, not instability.
5. `roots()` underpins `root_table()`, `plot_ar()`, `assess()` and `decompose_ar()` — understanding roots is understanding most of the package at once.
6. A pass from `assess()` is a structural precondition, not a guarantee of good performance on any particular event.
