---
title: "Using reach.postproc"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using reach.postproc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 5.5,
  warning = FALSE,
  message = FALSE
)

library(reach.postproc)
library(data.table)
library(ggplot2)

render_mermaid <- function(diagram) {
  if (!requireNamespace("DiagrammeR", quietly = TRUE)) {
    stop(
      paste0(
        "DiagrammeR is required to render Mermaid diagrams. ",
        "Install it with install.packages('DiagrammeR')."
      ),
      call. = FALSE
    )
  }

  DiagrammeR::mermaid(diagram)
}
```


# Purpose

This vignette provides a complete practical workflow. `reach.io` should import and standardise source data. `reach.rate` should apply any required level-flow conversion. `reach.postproc` starts from prepared series in the same domain and units.

# Choose a workflow

```{r task-flow, echo=FALSE}
render_mermaid("
flowchart TD
    A{What do you need to do?}
    B[Assess parameters]
    C[Project one error series]
    D[Evaluate an event archive]
    E[assess and roots]
    F[forecast_ar and apply_ar_update]
    G[align_forecast_series]
    H[fixed_lead_ar]
    I[score_lead_times and plot_lead_times]

    A --> B --> E
    A --> C --> F
    A --> D --> G --> H --> I
")
```

# Construct parameters

```{r workflow-parameters}
parameters <- default_ar_parameters()
parameters@coefficients
```

Explicit Deltares parameters:

```{r workflow-deltares}
deltares <- ar_parameters(
  coefficients = c(
    a_1 = 1.765,
    a_2 = -0.72625,
    a_3 = -0.040656
  ),
  sign_convention = "Deltares",
  label = "2024 defaults"
)
```

Equivalent standard signed-equation parameters:

```{r workflow-standard}
standard <- ar_parameters(
  coefficients = c(
    phi_1 = -1.765,
    phi_2 = 0.72625,
    phi_3 = 0.040656
  ),
  sign_convention = "Standard"
)

all.equal(deltares@coefficients, standard@coefficients)
```

# Inspect and assess roots

```{r workflow-roots}
root_information <- roots(parameters, time_step_minutes = 15)
root_table(root_information)
plot_ar(root_information, root_definition = "modal")
```

```{r workflow-assessment}
assessment <- assess(parameters)
assessment@summary
assessment_table(assessment)
```

Do not continue to operational evaluation with parameters that fail, unless the departure from the default criteria is documented and justified for the site.

# Forecast one error series

Initial errors are newest first.

```{r workflow-forecast}
error_forecast <- forecast_ar(
  parameters,
  initial_errors = c(0.15, 0.12, 0.10),
  steps = 96L,
  time_step_minutes = 15
)

forecast_series(error_forecast)
plot_ar(error_forecast)
```

Apply the errors to a simulated series:

```{r workflow-update}
simulated <- 0.8 + 1.2 * exp(
  -((forecast_series(error_forecast)$lead_time_hours - 12) / 4)^2
)

updated <- apply_ar_update(
  simulated = simulated,
  ar_error = forecast_series(error_forecast)$ar_error,
  lower_limit = 0,
  time = forecast_series(error_forecast)$lead_time_hours
)

updated
```

# Inspect root contributions

```{r workflow-decomposition}
components <- decompose_ar(
  parameters,
  initial_errors = c(0.15, 0.12, 0.10),
  steps = 96L,
  time_step_minutes = 15
)

ggplot(components, aes(lead_time_hours, contribution_real,
                       colour = factor(root_number))) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_line() +
  theme_minimal() +
  labs(colour = "Root", x = "Lead time, hours", y = "Contribution")
```

# Prepare reproducible event data

```{r workflow-data}
event_time <- seq(
  as.POSIXct("2024-11-23 00:00:00", tz = "UTC"),
  by = "15 min",
  length.out = 289L
)

simulated_values <- 0.55 + 1.35 * exp(
  -((seq_along(event_time) - 145) / 34)^2
)

observed_values <- simulated_values +
  0.12 +
  0.08 * sin(seq_along(event_time) / 12)

observed <- data.table(
  date_time = event_time,
  value = observed_values
)

simulated <- data.table(
  date_time = event_time,
  value = simulated_values
)
```

# Align the event

```{r workflow-align}
aligned <- align_forecast_series(
  observed = observed,
  simulated = simulated,
  interval_minutes = 15,
  timezone = "UTC",
  metadata = list(
    site_id = "example-site",
    event_id = "synthetic-event",
    measure = "level",
    units = "m"
  )
)

alignment_diagnostics(aligned)
aligned_series(aligned)
```

Stop if diagnostics show unexpected duplicates, gaps or irregular timesteps.

# Calculate fixed lead-time updates

```{r workflow-fixed-leads}
lead_results <- fixed_lead_ar(
  parameters = parameters,
  data = aligned,
  lead_times_minutes = c(30, 60, 90),
  time_step_minutes = 15,
  lower_limit = 0,
  method = "recurrence",
  metadata = list(
    site_id = "example-site",
    event_id = "synthetic-event"
  )
)

lead_time_series(lead_results)
```

# Score performance

```{r workflow-score}
scores <- score_lead_times(lead_results)
scores
```

Positive MAE or RMSE improvement means the updated series outperformed the simulation. Bias must be interpreted by distance from zero as well as sign.

# Plot results

```{r workflow-plot}
plot_lead_times(lead_results)
```

# Verify numerical equivalence

During package development, root projection should agree with recurrence projection within numerical tolerance.

```{r workflow-verification, eval=FALSE}
recurrence_results <- fixed_lead_ar(
  parameters,
  aligned,
  lead_times_minutes = c(30, 60, 90),
  method = "recurrence"
)

root_results <- fixed_lead_ar(
  parameters,
  aligned,
  lead_times_minutes = c(30, 60, 90),
  method = "roots"
)
```

# ARMA response analysis

```{r workflow-response}
response_data <- response_components(
  parameters,
  ma_parameters = c(-0.5, -0.25, -0.125),
  steps = 100L
)

ggplot(response_data, aes(step, response, colour = component)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_line() +
  theme_minimal()
```

# Export outputs

Calculation functions do not write files. Save only where the calling workflow requests it.

```{r workflow-export, eval=FALSE}
fwrite(
  lead_time_series(lead_results),
  "lead-time-series.csv"
)

fwrite(
  scores,
  "lead-time-scores.csv"
)

ggsave(
  "lead-time-plot.png",
  plot = plot_lead_times(lead_results),
  width = 300,
  height = 150,
  units = "mm",
  dpi = 300,
  bg = "white"
)
```

# Recommended sequence

1. Import with `reach.io`.
2. Convert with `reach.rate` if required.
3. Confirm common domain and units.
4. Align timestamps and inspect diagnostics.
5. Construct parameters with the correct sign convention.
6. Inspect roots and assess quality.
7. Calculate fixed lead-time updates.
8. Score and plot performance.
9. Compare methods during validation.
10. Export only required outputs.
